%

\chapter{Quantitative Input Data Usage}
\labch{quantitative-input-data-usage}

% \begin{itemize}
%   \item $\sup$ and $\inf$ operators;
%   \item $\length$ function;
%   \item abstract domain $\abstractdomain$
%   \item backward reachability state abstract semantics $\backwardsemanticsnoparam$.
% \end{itemize}

In this chapter, we present the quantitative framework used to measure the contributions of the input variables to the outcome of a program. First, we introduce a leading example of a landing-risk alarm system to illustrate the motivation of this chapter. Then, we present the quantitative framework and its properties. Since the framework depends on the underlying notion of impact, we show a few possible definitions of impact. Finally, we demonstrate the effectiveness of our approach against a set of potential applications.

\textit{Dans ce chapitre, nous présentons le cadre quantitatif utilisé pour mesurer les contributions des variables d'input à l'output d'un programme. Tout d'abord, nous introduisons un exemple principal d'un système d'alarme de risque d'atterrissage pour illustrer la motivation de ce chapitre. Ensuite, nous présentons le cadre quantitatif et ses propriétés. Étant donné que le cadre dépend de la notion d'impact, nous en montrons quelques définitions possibles. Enfin, nous démontrons l'efficacité notre approche contre des applications potentielles.}

\onlysectioncommands{\x,\y,\z,\lc,\exampleinput,\highlight,\inputa,\inputax,\inputay,\outputa,\inputb,\inputbx,\inputby,\outputb,\inputc,\inputcx,\inputcy,\outputc,\inputd,\inputdx,\inputdy,\outputd,\inpute,\inputex,\inputey,\outpute,\inputf,\inputfx,\inputfy,\outputf,\tracea,\traceax,\traceay,\traceb,\tracebx,\traceby,\tracec,\tracecx,\tracecy,\traced,\tracedx,\tracedy,\tracee,\traceex,\traceey,\tracef,\tracefx,\tracefy,\labelrotationangle,
\outputvaluea,\outputvalueb,\outputvaluec,\outputvalued,\outputvaluee,\outputvaluef}

\input{src/chapters/extensional/nfm24/overview.tex}
\input{src/chapters/extensional/nfm24/quantitative-input-data-usage}
% \input{src/chapters/extensional/nfm24/hierarchy}
\input{src/chapters/extensional/nfm24/static-analysis}
\input{src/chapters/extensional/nfm24/impatto}
