\chapter*{Acknowledgements and Ringraziamenti}
\addcontentsline{toc}{chapter}{Acknowledgements} % Add the preface to the table of contents as a chapter

First and foremost, I would like to thank my supervisor, Caterina Urban, for her guidance and support throughout these years. I am grateful for the opportunity you gave me, you have been a great mentor.
Although the initial challenges seemed insurmountable
, thanks to your patience and encouragement, I have been able to successfully complete this journey.
I would also like to thank Marco Campion for his help and remarkable advice. Thanks to you, I kept my motivation high.

I would like to thank Thomas Jensen and Isabella Mastroeni for reviewing this manuscript and providing valuable feedback. Thanks also to Bor-Yuh Evan Chang, Bernadette Charron-Bost, Matthieu Martel, Antoine Miné, and Corina Pǎsǎreanu for being part of my thesis committee and insightful discussions on my work.

I'm deeply grateful to have met Franco Raimondi during my London detour. You helped me navigate the challenges of my PhD and, generally, of my life.
I would like to also thank Corina Pǎsǎreanu for the endless research discussions and helping to build the foundations of my future career.

Thanks to the all the great members of the \emph{PVAR} Amazon Team that I had the pleasure to work, and have fun with.
Thanks to you, I have really enjoyed my time in London.
In particular, thanks to my boss Ilina, and mentors Evan, Franco and Daniel, without you my internship would not have been the same. Thanks also to all my London roommates and friends.

Je tiens à remercier les membres de l'équipe \emph{Antique} qui m'ont accompagné tout au long de ce voyage extraordinaire. Thanks to Arnold that helped me with the French sections of this manuscript, and to Đorđe for discussions on related work. I would also like to thank all the many students and researchers that I have met during my PhD at conferences, workshops, and summer schools for the interesting discussions and the great time spent together. To all the fellow PhD students of the \emph{Backseat Boys}:
Dimitri,
Đorđe,
Frederico,
Henry,
Ioana,
Jonáš,
Max,
Ramon, and
Sofia; tell the Chef that Paris is about 622 km far from Marktoberdorf.

And now is the time for friends I have met along the way. I would start with the \emph{Baiocchi}s:
Adam,
Albin,
Alessio,
Alessandro,
Charles,
Gabriele,
Ignacio,
Marco C., and
Patricio, thanks to all of you for the Parisian nights spent together.
I would like to thank the \emph{Biologists}:
Alberto,
Alessandra,
Alina,
Federico Crisp,
Federico,
Felix,
Gianmarco,
Julia,
Ilaria,
Ivan,
Héléna,
Leonardo,
Mari,
Marwaan,
Pol,
Romina,
Saverio, thanks for creating such a great group of friends.
Thanks to the climbers and \emph{Stone Seals}:
Arnold,
Alice,
Andrea,
Berenice,
Carol,
Francesco,
Lilit,
Lucas,
Mahiko,
Marita,
Pablo, and
Yuma,
for the many close-death and painful experiences.

Grazie agli amici Italiani che continuano ad essermi vicino.
Un grazie speciale a tutti quelli che sono potuti venire in persona alla mia difesa:
Alberto,
Alessandro,
Fabio,
Gianmarco,
Joel,
Luca,
Marcello,
Nico,
Riccardo, e
Tommaso.
Grazie anche a quelli che sono venuti a visitarmi a Parigi durante questi anni: Gianmarco, Luca, Ketto, e Tommaso.
Grazie agli amici dell'università Padovana:
Alessio,
Alessandro,
Andrea,
Ciprian,
Linpeng,
Paolo, e
Simone.

Ringrazio la mia famiglia per il loro supporto e fiducia in me. Grazie a Cecilia per avermi costantemente sopportato durante il periodo di scrittura della tesi e per aver transformato la mia visione di Parigi. I am grateful to Millo for being such a great climbing mentor and one of the closest friend I had here in Paris, I am really happy that our paths crossed. Last but not least, grazie a Francesco per avermi accompagnato lungo gli ultimi sei anni della mia vita e aver tirato fuori il meglio di me.
